Analysis paper

### 2021-08-27 analysis/analysis.Rmd
update ensemble plots

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
options(knitr.kable.NA = '')
library(here)
library(magrittr)
library(scoringutils)
library(knitr)
library(kableExtra)
library(data.table)
library(ggplot2)
library(ggridges)
library(RColorBrewer)
library(patchwork)
library(stringr)
library(covid.german.forecasts)
library(grates)
library(ggdist)
library(ggthemes)
library(scales)
library(dplyr)
library(tidyr)
library(forcats)

# define colors for deaths and cases
colcases <- brewer.pal(name = "Accent", n = 8)[-c(1, 4)]
coldeaths <- brewer.pal(name = "Accent", n = 8)[-4]

# colcases <- few_pal()(4)[-1]
# coldeaths <- few_pal()(4)

@@

distribution_plot_wis <- function(df, x = "interval_score") {
  df %>%
    ggplot(aes_string(y = "model", x = x, 
                      group = "model", fill = "model")) + 
    # geom_density_ridges(aes(fill = model, color = NA),
    #                     rel_min_height = 0.005,
    #                     scale = 0.9,
    #                     # jittered_points = TRUE,
    #                     # position = position_points_jitter(width = 0.05, height = 0),
    #                     point_size = 1.5,
    #                     point_shape = 21,
    #                     #point_shape = '|',
    #                     point_alpha = 1,
    #                     alpha = 0.3) +
    ggdist::stat_slab(alpha = 0.3, scale = 0.8, normalize = "panels") +
  # ggridges::geom_density_ridges(aes(color = NA), scale = 1.3) + 
    ggdist::stat_dots(aes(color = model)) + 
    ggdist::stat_pointinterval(point_interval = mean_qi, size = 3, shape = 16, fill = "black", .width = NA) + 
    ggdist::stat_pointinterval(point_interval = median_qi, size = 3, shape = 15, fill = "black", .width = NA) + 
        facet_wrap(~ target_type, ncol = 1, scales = "free") + 
    theme(plot.margin=unit(c(0,0.4,0,0),"cm")) +
    # geom_density_ridges(quantile_lines=TRUE,
    #                     quantile_fun=function(x,...) mean(x), 
    #                     linetype = "dotted", 
    #                     alpha = 0, 
    #                     scale = 0.9, 
    #                     rel_min_height = 0.01) +
    scale_fill_manual(values = coldeaths) + 
    scale_color_manual(values = coldeaths) + 
    # scale_color_manual(aesthetics = "point_color", 
    #                    values = brewer.pal(name = "Accent", n = 5)[1:5]) + 
    # theme(panel.background = element_rect(fill = "#e1f0fe")) +
    theme(legend.position = "bottom") + 
    labs(y = "Model", x = "Weighted interval score") + 
  scale_x_continuous(labels = scales::label_number(suffix = "k", 
                                                   scale = 1e-3, 
                                                   accuracy = NULL)) + 
    theme_minimal()
}

for (i in 1:4) {
  p_combined_wis <- distribution_plot_wis(df = scores[horizon == i], 
                                          x = "interval_score") +
    facet_wrap(~ target_type, ncol = 1, scales = "free") + 
    theme(plot.margin=unit(c(0,0.4,0,0),"cm"))

@@

  pl_wis <- wrap_plots(p_combined_wis, p_wis, widths = c(1, 2)) / p_phases_wis + 
    plot_layout(guides = "collect") & 
    plot_annotation(tag_levels = 'A') &
    theme(legend.position = 'bottom')

  ggsave(here("analysis", "plots", 
              paste0("distribution_scores_wis-", i, ".png")), 
         pl_wis, height = 11, width = 9)

@@

p_combined_ae <- distribution_plot_wis(df = scores, 
                                        x = "aem") +
  facet_wrap(~ target_type, ncol = 1, scales = "free_x")

# p <- distribution_plot(df = scores[model != "Baseline"]) + 
#   theme(axis.title.y=element_blank(),
#         axis.text.y=element_blank(), 
#         axis.ticks.y = element_blank()) + 
#   facet_wrap(~ location_target, ncol = 2)

@@

aggregate_performance <- function(data, 
                                  x = "horizon",
                                  xlabel = "Forecast horizon",
                                  output_dir = here("analysis", "plots"), 
                                  plotname,
                                  relative = FALSE,
                                  facet_formula = ~ target_type) {

  scores <- eval_forecasts(data, 
                           summarise_by = c("model", "target_type", x), 
                           quantiles = 0.5, 
                           sd = TRUE)

  setnames(scores, old = "model", new = "Model")

  plot_helper <- function(tmp = scores, xvar = x, y, rel = relative) {
    if (rel) {
      tmp <- tmp %>%
        arrange(target_type, get(x)) %>%
        group_by(target_type, get(x)) %>%
        mutate_at(y, ~ . - .[Model == "Hub-ensemble"])
    }

    tmp %>%
      ggplot(aes_string(x = xvar, y = y, group = "Model", color = "Model")) +
      geom_point() +
      geom_line() +
      facet_wrap(facet_formula, scales = "free_y") +
      theme_minimal() +
      theme(legend.position = "bottom") +
      scale_color_manual(values = coldeaths)
  }

  # mean performance
  p_mean <- 
    plot_helper(scores, y = "interval_score") + 
    labs(y = "mean WIS", x = NULL)

    # absolute error of performance
  p_aem <-
    plot_helper(scores, y = "aem") + 
    labs(y = "Absolute error", x = NULL)

  # sd of performance
  p_wis_sd <- 
    plot_helper(scores, y = "interval_score_sd") + 
    labs(y = "WIS - sd", x = NULL)

  # sharpness of performance
  p_sharpness <- 
    plot_helper(scores, y = "sharpness") + 
    labs(y = "Sharpness", x = NULL)

  # bias
  p_bias <- 
    plot_helper(scores, y = "bias") + 
    labs(y = "Bias", x = NULL) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    expand_limits(y=c(-0.2, 0.2))

  data <- filtered_data[model %in% regular_models]

  scores <- eval_forecasts(data, 
                           summarise_by = c("model", "target_type", "horizon"), 
                           quantiles = 0.5, 
                           sd = TRUE)

  setnames(scores, old = "model", new = "Model")

  component_plot <- function(scores, type) {

    if (type == "case") {
      vals <- colcases
    } else {
      vals <- coldeaths
    }

    scores %>%
      select(Model, target_type, horizon, sharpness, interval_score,
             underprediction, overprediction) %>%
      filter(target_type == type) %>%
      pivot_longer(names_to = "WIS components",
                   cols = c(sharpness, overprediction, underprediction)) %>%
      mutate(`WIS components` = fct_relevel(`WIS components`, 
                                            c("sharpness", "underprediction", 
                                              "overprediction"))) %>%
      group_by(target_type, horizon) %>%
      arrange(interval_score) %>%
      ggplot(aes(x = reorder(Model, (interval_score)), y = value, fill = Model, 
                 group = Model, alpha = `WIS components`)) + 
      scale_alpha_manual(values = c(1, 0.1, 0.6)) +
      scale_fill_manual(values = vals) + 
      geom_bar(stat = "identity", position = "stack",
               color = "black",
               size = 0.1,
               width = 0.8) +
      facet_wrap(~ horizon, ncol = 4,
                 strip.position="bottom") + 
      theme_minimal() + 
      theme(axis.text.x = element_blank(), 
            plot.title = element_text(hjust = 0.5, 
                                      size = 8), 
            legend.position = c(0.12, 0.83)) + 
      labs(x = NULL, title = type, y = "WIS") + 
      guides(fill = "none")
  }

  component_case <- component_plot(scores, "case")
  component_death <- component_plot(scores, "death")
  components <- component_case + component_death

  # coverage
  scores <- eval_forecasts(data, 
                           summarise_by = c("model", "target_type", x, "range"), 
                           quantiles = 0.5, 
                           sd = TRUE)

  setnames(scores, old = "model", new = "Model")

  p_cov50 <- scores[range %in% c(50)] %>%
    plot_helper(y = "coverage", rel = FALSE) + 
    geom_hline(yintercept = 0.5, linetype = "dashed") +
    labs(y = "Coverage - 50% PI") + 
    expand_limits(y=c(0, 1)) + 
    scale_y_continuous(labels = scales::percent)

  p_cov90 <- scores[range %in% c(90)] %>%
    plot_helper(y = "coverage", rel = FALSE) + 
    geom_hline(yintercept = 0.9, linetype = "dashed") +
    labs(y = "Coverage - 90% PI") + 
    expand_limits(y=c(0, 1)) + 
    scale_y_continuous(labels = scales::percent)

@@

filtered_data[model %in% regular_models & horizon == 2] %>%
  aggregate_performance(plotname = "aggregate-performance-2-weeks-locations-all", 
                        x = "location_name", 
                        xlabel = NULL)

filtered_data[model %in% ensemble_models[!grepl("mean", ensemble_models)]] %>%
  aggregate_performance(plotname = "aggregate-performance-rel-ensemble", 
                        relative = TRUE)

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, cache = TRUE)
options(knitr.kable.NA = '')
library(here)
library(magrittr)
library(scoringutils)
library(knitr)
library(kableExtra)
library(data.table)
library(ggplot2)
library(ggridges)
library(RColorBrewer)
library(patchwork)
library(stringr)
library(covid.german.forecasts)
library(grates)
library(ggdist)
library(ggthemes)
library(scales)
library(dplyr)
library(tidyr)
library(forcats)

# define colors for deaths and cases
colcases <- brewer.pal(name = "Accent", n = 8)[-c(1, 4)]
coldeaths <- brewer.pal(name = "Accent", n = 8)[-4]
colensemble <- c("#FDC086", brewer.pal(name = "Set2", n = 8))

@@

distribution_plot_wis <- function(df, x = "interval_score", ranks = FALSE) {

  scale_fn <- function(x) {
    ifelse(x >= 5000,
           paste0(x / 1000, "k"), x)
  }

  xlabel = "Weighted interval score"
  dotsize = 1

  if (ranks) {
    df <- df %>%
      group_by(location_name, target_type, horizon, forecast_date) %>%
      arrange(location_name, target_type, horizon, forecast_date, interval_score) %>%
      mutate(!!x := rank(get(x)))

    xlabel <- "Rank in terms of WIS"
    dotsize = 0.1

  }
  + runif(length(get(x)), min = -0.1, max = 0.1)

  df %>%
    rename(Model = model) %>%
    ggplot(aes_string(y = "Model", x = x, 
                      group = "Model", fill = "Model")) + 
    ggdist::stat_slab(alpha = 0.3, scale = 0.8, normalize = "panels") +
    ggdist::stat_dots(aes(color = Model),
                      dotsize = dotsize) + 
    ggdist::stat_pointinterval(point_interval = mean_qi, size = 3, 
                               shape = 16, fill = "black", .width = NA) + 
    ggdist::stat_pointinterval(point_interval = median_qi, size = 3, 
                               shape = 15, fill = "black", .width = NA) + 
        facet_wrap(~ target_type, ncol = 1, scales = "free") + 
    theme(plot.margin=unit(c(0,0.4,0,0),"cm")) +
    scale_fill_manual(values = coldeaths) + 
    scale_color_manual(values = coldeaths) + 
    theme(legend.position = "bottom") + 
    labs(y = "", x = xlabel) + 
  scale_x_continuous(labels = scale_fn) + 
    theme_minimal()
}

for (i in 1:4) {
  p_combined_wis <- distribution_plot_wis(df = scores[horizon == i], 
                                          x = "interval_score") +
    facet_wrap(~ target_type, ncol = 1, scales = "free") + 
    theme(plot.margin=unit(c(0,0.4,0,0),"cm")) + 
    theme(legend.position = "none") + 
    guides(shape = "none", colour = "none", fill = "none")

@@

  # pl_wis <- wrap_plots(p_combined_wis, p_wis, widths = c(1, 2)) / p_phases_wis + 
  pl_wis <- p_combined_wis + p_wis +
    plot_layout(guides = "collect", 
                widths = c(1, 2)) &
    plot_annotation(tag_levels = 'A')  &
    theme(legend.position = 'bottom', 
          legend.box="vertical", legend.margin=margin())

  ggsave(here("analysis", "plots", 
              paste0("distribution_scores_wis-", i, ".png")), 
         plot = pl_wis, height = 6, width = 9)

@@

# plot with rank distributions
for (i in 1:4) {
  p_combined_wis <- distribution_plot_wis(df = scores[horizon == i], 
                                          x = "interval_score", 
                                          ranks = TRUE) +
    facet_wrap(~ target_type, ncol = 1, scales = "free") + 
    theme(plot.margin=unit(c(0,0.4,0,0),"cm")) + 
    theme(legend.position = "none") + 
    guides(shape = "none", colour = "none", fill = "none")

  p_wis <- distribution_plot_wis(df = scores[horizon == i], 
                                 ranks = TRUE) + 
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(), 
        axis.ticks.y = element_blank()) + 
  facet_wrap(~ location_target, ncol = 2, scale = "free")

  # pl_wis <- wrap_plots(p_combined_wis, p_wis, widths = c(1, 2)) / p_phases_wis + 
  pl_wis <- p_combined_wis + p_wis +
    plot_layout(guides = "collect", 
                widths = c(1, 2)) &
    plot_annotation(tag_levels = 'A')  &
    theme(legend.position = 'bottom', 
          legend.box="vertical", legend.margin=margin())

  ggsave(here("analysis", "plots", 
              paste0("distribution_scores_wis-", i, "-ranks.png")), 
         plot = pl_wis, height = 6, width = 9)

# same plot for absolute error
p_combined_ae <- distribution_plot_wis(df = scores, 
                                        x = "aem") +
  facet_wrap(~ target_type, ncol = 1, scales = "free_x")

@@

# function to make plot with mean of a given metric
plot_helper <- function(tmp = scores, xvar = x, y, rel = relative, 
                        facet_formula, colour) {
  if (rel) {
    tmp <- tmp %>%
      arrange(target_type, !!sym(xvar)) %>%
      group_by(target_type, !!sym(xvar)) %>%
      mutate_at(y, ~ . / .[Model == "Hub-ensemble"])
  }

  tmp %>%
    ggplot(aes_string(x = xvar, y = y, group = "Model", color = "Model")) +
    geom_point(position = position_dodge(width = 0.1)) +
    geom_line(position = position_dodge(width = 0.1)) +
    facet_wrap(facet_formula, scales = "free_y") +
    theme_minimal() +
    theme(legend.position = "bottom") +
    scale_color_manual(values = colour)
}

# function for creating a component plot
component_plot <- function(scores, type, x = x, rel = relative,
                           facet_formula, vals = NULL) {

  if (type == "case" & is.null(vals)) {
    vals <- colcases
  } else if (type == "death" & is.null(vals)){
    vals <- coldeaths
  } else if (type == "case") {
    # filter out ensembles not needed
    scores <- scores %>%
      filter(!(Model %in% c("Hub-ensemble-with-all", "Hub-ensemble-with-convolution")))
    vals <- vals[-c(3, 4)]
  }

  scores <- scores %>%
    select(Model, target_type, sharpness, interval_score,
           underprediction, overprediction, !!x) %>%
    filter(target_type == type) %>%
    pivot_longer(names_to = "WIS components",
                 cols = c(sharpness, overprediction, underprediction)) %>%
    mutate(`WIS components` = fct_relevel(`WIS components`, 
                                          c("sharpness", "underprediction", 
                                            "overprediction")))

  if (rel) {
    scores <- scores %>%
      arrange(!!sym(x), target_type, interval_score) %>%
      group_by(!!sym(x), target_type) %>%
      mutate_at("value", ~ . / interval_score[Model == "Hub-ensemble"])
  }

  p <- scores %>%
    ggplot(aes(x = reorder(Model, (interval_score)), y = value, fill = Model, 
               group = Model, alpha = `WIS components`))

  if (rel) {
    p <- p + 
      geom_hline(yintercept = 1, linetype = "dashed")
  }

  p <- p + 
    scale_alpha_manual(values = c(1, 0.1, 0.6)) +
    scale_fill_manual(values = vals) + 
    geom_bar(stat = "identity", position = "stack",
             color = "black",
             size = 0.1,
             width = 0.8) +
    facet_wrap(as.formula(paste("~", x)), ncol = 4,
               strip.position="bottom") + 
    theme_minimal() + 
    theme(axis.text.x = element_blank(), 
          plot.title = element_text(hjust = 0.5, 
                                    size = 8), 
          legend.position = c(0.12, 0.83)) + 
    labs(x = NULL, title = type, y = "WIS") + 
    guides(fill = "none")

  return(p)
}

score_helper <- function(data, summarise_by) {
  scores <- eval_forecasts(data, 
                           summarise_by = summarise_by, 
                           quantiles = 0.5, 
                           sd = TRUE) %>%
    filter(!(model %in% c("Hub-ensemble-with-all", 
                          "Hub-ensemble-with-convolution") & 
               target_type == "case"))

  setnames(scores, old = "model", new = "Model")
}

aggregate_performance <- function(data, 
                                  x = "horizon",
                                  xlabel = "Forecast horizon",
                                  output_dir = here("analysis", "plots"), 
                                  plotname,
                                  relative = FALSE,
                                  facet_formula = ~ target_type, 
                                  colour_scheme = c("normal", "ensemble"),
                                  vals = NULL) {

  # create scores
  scores <- score_helper(data, c("model", "target_type", x))

  if (colour_scheme[1] == "normal") {
    vals <- coldeaths
  } else {
    vals <- colensemble
  }

  # mean performance
  p_mean <- 
    plot_helper(scores, y = "interval_score", rel = relative, 
                facet_formula = facet_formula, colour = vals) + 
    labs(y = "mean WIS", x = NULL)

    # absolute error of performance
  p_aem <-
    plot_helper(scores, y = "aem", rel = relative, 
                facet_formula = facet_formula, colour = vals) + 
    labs(y = "Absolute error", x = NULL)

  # sd of performance
  p_wis_sd <- 
    plot_helper(scores, y = "interval_score_sd", rel = relative, 
                facet_formula = facet_formula, colour = vals) + 
    labs(y = "WIS - sd", x = NULL)

  # sharpness of performance
  p_sharpness <- 
    plot_helper(scores, y = "sharpness", rel = relative, 
                facet_formula = facet_formula, colour = vals) + 
    labs(y = "Sharpness", x = NULL)

  # bias plot
  bias_limits <- max(abs(scores$bias)) * c(-1.1, 1.1) %>% 
    round(2)
  p_bias <- 
    plot_helper(scores, y = "bias", rel = FALSE, colour = vals, 
                facet_formula = facet_formula) + 
    labs(y = "Bias", x = NULL) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    expand_limits(y=bias_limits)

  # plot with wis components
  component_case <- component_plot(scores, "case", x = x, vals = vals, 
                                   facet_formula = facet_formula, rel = relative)
  component_death <- component_plot(scores, "death", x = x, vals = vals, 
                                    facet_formula = facet_formula, rel = relative)
  components <- component_case + component_death

  # coverage plots 
  scores <- score_helper(data,  c("model", "target_type", x, "range"))

  if (relative) {
    scores <- scores %>%
      arrange(target_type, range, !!sym(x)) %>%
      group_by(target_type, range, !!sym(x)) %>%
      mutate_at("coverage", ~ . - .[Model == "Hub-ensemble"])
  }

  p_cov50 <- scores %>%
    filter(range == c(50)) %>%
    plot_helper(y = "coverage", rel = FALSE, , facet_formula = facet_formula, 
                colour = vals) +
    labs(y = "Coverage - 50% PI") + 
    scale_y_continuous(labels = scales::percent)

  if (!relative) {
    p_cov50 <- p_cov50 + 
      expand_limits(y=c(0, 1)) + 
      geom_hline(yintercept = 0.5, linetype = "dashed") + 
      labs(y = "Diff. cov. - 50% PI", x = "") + 
      scale_y_continuous()
    }

  p_cov90 <-  scores %>%
    filter(range == c(90)) %>%
    plot_helper(y = "coverage", rel = FALSE, facet_formula = facet_formula, 
                colour = vals) + 
    labs(y = "Coverage - 90% PI", x = "") + 
    scale_y_continuous(labels = scales::percent)

    if (!relative) {
      p_cov90 <- p_cov90 + 
        expand_limits(y=c(0, 1)) + 
        geom_hline(yintercept = 0.9, linetype = "dashed") + 
        labs(y = "Diff. cov. - 90% PI") + 
        scale_y_continuous()
    }

@@

# plot for locations
filtered_data[model %in% regular_models & horizon == 2] %>%
  aggregate_performance(plotname = "aggregate-performance-2-weeks-locations-all", 
                        x = "location_name", 
                        xlabel = NULL)

# relative plots for locations
filtered_data[model %in% regular_models & horizon == 2] %>%
  aggregate_performance(plotname = "aggregate-performance-2-weeks-locations-all-rel", 
                        x = "location_name", 
                        xlabel = NULL, relative = TRUE)

# plot for ensemble
filtered_data[model %in% ensemble_models[!grepl("mean", ensemble_models)]] %>%
  # filter(!(model == "Hub-ensemble-with-all" & target_type == "case")) %>%
  aggregate_performance(plotname = "aggregate-performance-rel-ensemble", 
                        relative = TRUE, colour_scheme = "ensemble")
